---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# lfl - Linguistic Fuzzy Logic

<!-- badges: start -->
[![CRAN status](https://www.r-pkg.org/badges/version/lfl)](https://CRAN.R-project.org/package=lfl)
[![codecov](https://codecov.io/gh/beerda/lfl/branch/master/graph/badge.svg)](https://app.codecov.io/gh/beerda/lfl)
[![Downloads](https://cranlogs.r-pkg.org/badges/grand-total/lfl)](https://cran.r-project.org/package=lfl)
<!-- badges: end -->

The `lfl` package for [R statistical computing environment](https://www.r-project.org/) provides various algorithms related to **linguistic fuzzy logic**: mining for linguistic fuzzy association rules, composition of fuzzy relations, performing perception-based logical deduction (PbLD), and forecasting time-series using fuzzy rule-based ensemble (FRBE). The package also contains basic fuzzy-related algebraic functions capable of handling missing values in different styles (Bochvar, Sobocinski, Kleene etc.), computation of Sugeno integrals and fuzzy transform.

## Key Features

- **Fuzzy Algebras**: Support for various t-norms (Gödel, Goguen, Łukasiewicz) and corresponding t-conorms, residua, bi-residua, and negations
- **Missing Value Handling**: Extensions for handling NAs using Kleene, Bochvar, Sobocinski, Dragonfly, and other approaches
- **Linguistic Expressions**: Tools for working with evaluative linguistic expressions (e.g., "very small", "roughly medium", "extremely big")
- **Fuzzy Relations**: Composition of fuzzy relations with various extensions
- **Association Rules**: Mining for linguistic fuzzy association rules from data
- **Perception-based Logical Deduction (PbLD)**: Inference method specifically designed for linguistic fuzzy logic
- **Fuzzy Rule-Based Ensemble (FRBE)**: Time series forecasting using ensemble of forecasting methods
- **Fuzzy Transform**: Computing fuzzy transform and its inverse
- **Sugeno Integral**: Computation of Sugeno integrals

## Installation

To install the stable version from CRAN, type the following command within your R session:

```r
install.packages("lfl")
```

You can also install the development version of `lfl` from [GitHub](https://github.com/) with:

```r
install.packages("devtools")
devtools::install_github("beerda/lfl")
```

To start using the package, load it to the R session with:

```{r}
library(lfl)
```

## Quick Examples

### Creating Fuzzy Algebras

```{r}
# Create a Gödel algebra
a <- algebra("goedel")

# Compute fuzzy operations
a$n(c(0.5, 0.8, 0, 1))  # negation
a$t(c(0.8, 0.3), c(0.2, 1), c(1, 0))  # t-norm (conjunction)
a$s(c(0.8, 0.3), c(0.2, 1))  # t-conorm (disjunction)
a$r(c(0.8, 0.3), c(0.2, 1))  # residuum (implication)
```

### Composition of Fuzzy Relations

```{r}
# Define fuzzy relations
R <- matrix(c(0.9, 0, 0, 0,
              1, 0.9, 0.8, 0,
              0.8, 0.8, 0.9, 1,
              0, 0.1, 0, 0.9), nrow=4)

S <- matrix(c(1, 0.9, 0, 1,
              1, 0.2, 1, 0,
              0.1, 0.9, 0, 0.7,
              0.9, 0, 1, 0.1,
              0, 1, 1, 0.9), nrow=4)

# Compute composition
compose(R, S, alg=algebra("lukasiewicz"), type="basic")
```

### Evaluative Linguistic Expressions

```{r}
# Create context for linguistic expressions
context <- ctx3(low=0, center=50, high=100)

# Create linguistic expressions
small <- lcut(1:100, context, atomic="sm")
medium <- lcut(1:100, context, atomic="me")
big <- lcut(1:100, context, atomic="bi")
```

## Documentation

For complete documentation, please visit the [package website](https://beerda.github.io/lfl/) or read the vignettes:

- **Introduction to lfl**: Overview of the package features
- **Fuzzy Algebras**: Detailed guide on fuzzy algebras and handling missing values
- **Composition of Fuzzy Relations**: Working with fuzzy relation compositions
- **Perception-based Logical Deduction**: Using PbLD for fuzzy inference
- **Fuzzy Rule-Based Ensemble**: Time series forecasting with FRBE

You can also access the original academic article describing the package:
[Burda & Štěpnička (2017). lfl: An R Package for Linguistic Fuzzy Logic](https://github.com/beerda/lfl/blob/master/vignettes/main.pdf).

## Contributing

Contributions, suggestions, and bug reports are welcome. Please submit [issues](https://github.com/beerda/lfl/issues/) on [GitHub](https://github.com/).

## Citation

If you use this package in your research, please cite:

```
Burda, M., & Štěpnička, M. (2017). lfl: An R Package for Linguistic Fuzzy Logic.
```

## License

This package is free software: you can redistribute it and/or modify it under the terms of the GNU General Public License as published by the Free Software Foundation, either version 3 of the License, or (at your option) any later version.
