---
output:
  github_document:
    html_preview: false
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

[![Travis Build Status](https://travis-ci.org/beerda/lfl.svg?branch=master)](https://travis-ci.org/beerda/lfl)
[![AppVeyor Build Status](https://ci.appveyor.com/api/projects/status/github/beerda/lfl?branch=master&svg=true)](https://ci.appveyor.com/project/beerda/lfl)
[![codecov](https://codecov.io/gh/beerda/lfl/branch/master/graph/badge.svg)](https://codecov.io/gh/beerda/lfl)
[![CRAN_Status_Badge](http://www.r-pkg.org/badges/version/lfl)](https://cran.r-project.org/package=lfl)


# lfl

The lfl package provides various algorithms related to linguistic fuzzy logic: mining for linguistic fuzzy association
rules, composition of fuzzy relations, performing  perception-based logical deduction (PbLD),  and forecasting
time-series using fuzzy rule-based ensemble (FRBE).

## Installation

To install the stable version from CRAN, simply issue the following command within your R session:

```{r cran-installation, eval = FALSE}
install.packages("lfl")
```

If you want to install the development version instead, type:

```{r gh-installation, eval = FALSE}
install.packages("devtools")
devtools::install_github("beerda/lfl")
```
